Evolution Reinforces Cooperation with the Emergence of Self-Recognition Mechanisms: an Empirical Study of Strategies in the Moran process for the Iterated Prisoner's Dilemma

### 2018-07-17 main.tex
Remove all images.
Add more details about the structured strategies.
Have added a general description of FSMs but also pointed (in various
places) at the ml paper.

% Header and Footer with logo
\usepackage{lastpage,fancyhdr}
\usepackage{graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

@@

This manuscript provides a detailed numerical analysis of agent-based simulations
of \textbf{164} complex and adaptive strategies for the
IPD\@. This is made possible by the Axelrod library \cite{axelrodproject}, an
effort to provide software for reproducible research for the IPD\@. The library
now contains over 186 parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. Not all strategies have been considered for this study: excluded
are those that make use of knowledge of the number of turns in a match
and others that have a high
computational run time. The large number of strategies are available thanks to
the open source nature of the project with over 50 contributors from around the
world, made by programmers and researchers \cite{Knight2016}. Three of the considered
strategies are finite state machines trained specifically for Moran processes
(described further in the Methods section).

@@

A number of these strategies have been trained with reinforcement learning
algorithms prior to this study and not specifically for the Moran process.

@@

Apart from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. They were trained to
win IPD tournaments by maximizing their mean total payoffs against a variety
of opponents. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.
There has also been some work on strategies using an evolutionary algorithm in real time:
in~\cite{Gaudesi2016}
an evolutionary algorithm is used to build a model of the opponent and attempt
to exploit any potential weakness. In this work all strategies resulting from
evolutionary algorithms are pre-trained.

@@

All of the training code is archived at \cite{marc_harper_2017_824264}. This
software is (similarly to the Axelrod library) available on
GitHub\footnote{\url{https://github.com/Axelrod-Python/axelrod-dojo}}
with documentation to
train new strategies easily. Training
typically takes less than 100 generations and can be completed within several
hours on commodity hardware.

There are three further strategies trained specifically for this study; Trained
FSM 1, 2, and 3 (TF1 - TF3). These are based on finite state machines of 16, 16,
and 8 states respectively. These are shown in Fig~\ref{fig:tf1},~\ref{fig:tf2}
and~\ref{fig:tf3}, using the notation common in the literature where
\(A_1/A_2\) is the action of the opponent \(A_1\) and the response of the player
\(A_2\) as well as arrows corresponding to changes of state.

>>>>

% Header and Footer with logo
\usepackage{lastpage,fancyhdr}
\usepackage[draft]{graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

@@

This manuscript provides a detailed numerical analysis of agent-based simulations
of \textbf{164} complex and adaptive strategies for the
IPD\@. This is made possible by the Axelrod library~\cite{axelrodproject}, an
effort to provide software for reproducible research for the IPD\@. The library
now contains over 186 parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. Not all strategies have been considered for this study: excluded
are those that make use of knowledge of the number of turns in a match
and others that have a high
computational run time. The large number of strategies are available thanks to
the open source nature of the project with over 50 contributors from around the
world, made by programmers and researchers~\cite{Knight2016}. Three of the considered
strategies are finite state machines trained specifically for Moran processes
(described further in the Methods section).

@@

Using families of strategies that depend on given parameters it is possible to
find specific parameters through a training process called reinforcement
learning. A detailed description of the various types considered is given
in~\cite{Harper2017}.

A number of these strategies have been trained this way (see~\cite{Harper2017})
prior to this study and not specifically for the Moran process.

@@

Apart from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents~\cite{affenzeller2009genetic}. They were trained to
win IPD tournaments by maximizing their mean total payoffs against a variety
of opponents. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature~\cite{Ashlock2006}.
There has also been some work on strategies using an evolutionary algorithm in real time:
in~\cite{Gaudesi2016}
an evolutionary algorithm is used to build a model of the opponent and attempt
to exploit any potential weakness. In this work all strategies resulting from
evolutionary algorithms are pre-trained.

@@

All of the training code is archived at~\cite{marc_harper_2017_824264}. This
software is (similarly to the Axelrod library) available on
GitHub\footnote{\url{https://github.com/Axelrod-Python/axelrod-dojo}}
with documentation to
train new strategies easily. Training
typically takes less than 100 generations and can be completed within several
hours on commodity hardware.

One particular family of strategies that has been studied in the literature are
called finite state machines. These mathematical models consist of states and
and responses to actions from one state to another. In the context of the IPD, a
finite state machine, is a mapping from states and opponent actions to states
and an action. For further details on this type of strategy, the reader is
referred to~\cite{Ashlock2006, Harper2017}.

There are three further strategies trained specifically for this
study; Trained FSM 1, 2, and 3 (TF1 - TF3). These are finite state
machines of 16, 16, and 8 states respectively.  These are shown in
Fig~\ref{fig:tf1},~\ref{fig:tf2} and~\ref{fig:tf3}, using the notation common in
the literature where \(A_1/A_2\) is the action of the opponent \(A_1\) and the
response of the player \(A_2\) as well as arrows corresponding to changes of
state.
